Stagnating State Support for Higher Education and Its Effect on Faculty Composition

### 2022-08-15 text/paper.tex
rename.
adjust script and README.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Title page
% Author
\author{Senan Hogan-Hennessy\thanks{
    Cornell University, Department of Economics.}
    \thanks{I thank Professor Evan Riehl for guidance on this project, Michael Lovenheim, Ronald Ehrenberg, Douglas Miller, and Francine Blau for various comments on the project's direction,
    as well as seminar participants at Cornell University (Autumn 202) for helpful
    discussion.
    This project's Github repository, with materials and (publicly available) data for replication, is available at 
    \url{https://github.com/shoganhennessy/finances-faculty-composition}.
    Any comments or suggestions may be sent to me at \href{mailto:seh325@cornell.edu}{\nolinkurl{seh325@cornell.edu}}, or raised as an issue on the Github project.}}
% Title
\title{Stagnating State Support for Higher Education and Its Effect on Faculty Composition}
\date{\today}
\rhead{\today}
\lhead{State Support for Higher Education and Faculty Composition}
% Begin
\begin{document}
\clearpage \maketitle
\thispagestyle{empty}
% Abstract
\begin{abstract}
\noindent
State support for higher education has stagnated on a per student level the last thirty years, while public universities have reduced the number of tenure-track or with tenured professors per student, and increased their employment of contingent lecturers.
This analysis relates these changes in the faculty composition at public universities to the fall in per student university revenues from state appropriations.
A shift-share instrumental variables approaches addresses endogeneity in decisions for state appropriations to higher education, by exploiting the yearly change in state support with an institution's reliance on that support.
A decrease in state funding of 10\%, via a shock to state appropriations, decreases the number of assistant professors per student at a public university by 1.4\% and full professors by 1.2\%, yet increases the number of lecturers per student by 4.4\%.
Analysis of all the professors hired at Illinois public universities 2011-2021 shows that incumbent professors are not affected by the changes in state support, implying that these changes in faculty composition arose by lower rate of hiring of tenure-track and tenured professors at public universities.
These results show one of the long-term effects of stagnating state support for higher education, and raises questions about the direction that public education heads as these financial headwinds show no sign of dissipating.

@@

\cite{NBERw23736} first study the effect of (changes in) public universities' finances on student outcomes by isolating changes in public universities' expenditures on student instruction, addressing endogeneity of expenditures decisions by instrumenting expenditures with state appropriation shocks and limits on tuition price.
\cite{NBERw23736} find that increases in public university spending (via appropriation shocks) increases enrolment and degree completion among students, but corresponding changes in attendance price (via limits on tuition price) do not have an effect.
\cite{chakrabarti2018effect,NBERw27885} use the same instrument for public university revenues at the state level, combined with rich information on students' individual-level outcomes, to show that increases in state appropriations lead to degree lower completion time and later-life student debt.
\cite{bound2019public} use another variant of the same approach to show that state-level higher education spending cuts induced public universities to shift toward tuition as their primary source of revenue, and shift institutional resources towards students who pay higher tuition rates.
In a similar vein, \cite{bound2007cohort} document variation in per student state funding for higher education resulting from yearly changes in a state's student-age population; they find that lower per student funding lead to lower higher education completion rate among students.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Title page
% Author
\author{Senan Hogan-Hennessy\thanks{
    Cornell University, Department of Economics.}
    \thanks{I thank Professor Evan Riehl for guidance on this project, Michael Lovenheim, Ronald Ehrenberg, Douglas Miller, and Francine Blau for various comments on the project's direction,
    as well as seminar participants at Cornell University (Autumn 202) for helpful
    discussion.
    This project's Github repository, with materials and (publicly available) data for replication, is available at 
    \url{https://github.com/shoganhennessy/state-faculty-composition}.
    Any comments or suggestions may be sent to me at \href{mailto:seh325@cornell.edu}{\nolinkurl{seh325@cornell.edu}}, or raised as an issue on the Github project.}}
% Title
\title{Stagnating State Support for Higher Education and Its Effect on Faculty Composition}
\date{\today}
\rhead{\today}
\lhead{State Support for Higher Education and Faculty Composition}
% Begin
\begin{document}
\clearpage \maketitle
\thispagestyle{empty}
% Abstract
\begin{abstract}
\noindent
State support for higher education has stagnated on a per student level the last thirty years, while public universities have reduced the number of tenure-track or with tenured professors per student, and increased their employment of contingent lecturers.
This analysis relates these changes in the faculty composition at public universities to the fall in per student university revenues from state appropriations.
A shift-share instrumental variables approaches addresses endogeneity in decisions for state appropriations to higher education, by exploiting the yearly change in state support with an institution's reliance on that support.
A decrease in state funding of 10\%, via a shock to state appropriations, decreases the number of assistant professors per student at a public university by 1.4\% and full professors by 1.2\%, yet increases the number of lecturers per student by 4.4\%.
Analysis of all the professors hired at Illinois public universities 2011-2021 shows that incumbent professors are not affected by the changes in state support, implying that these changes in faculty composition arose by lower rate of hiring of tenure-track and tenured professors at public universities.
These results show one of the long-term effects of stagnating state support for higher education, and raises questions about the direction that public education heads as these financial headwinds show no sign of dissipating.

@@

\cite{NBERw23736} first study the effect of (changes in) public universities' finances on student outcomes by isolating changes in public universities' expenditures on student instruction, addressing endogeneity of expenditures decisions by instrumenting expenditures with state appropriation shocks and limits on tuition price.
\cite{NBERw23736} find that increases in public university spending (via appropriation shocks) increases enrolment and degree completion among students, but corresponding changes in attendance price (via limits on tuition price) do not have an effect.\footnote{
    \cite{miller2022making} further analyse the effects of falls in university revenues, by finding that reductions in public universities tuition prices (suggestively motivated by falling state funding) leads to reductions in provided financial aid.
}
\cite{chakrabarti2018effect,NBERw27885} use the same instrument for public university revenues at the state level, combined with rich information on students' individual-level outcomes, to show that increases in state appropriations lead to degree lower completion time and later-life student debt.
\cite{bound2019public} use another variant of the same approach to show that state-level higher education spending cuts induced public universities to shift toward tuition as their primary source of revenue, and shift institutional resources towards students who pay higher tuition rates.
In a similar vein, \cite{bound2007cohort} document variation in per student state funding for higher education resulting from yearly changes in a state's student-age population; they find that lower per student funding lead to lower higher education completion rate among students.
